---
title: "ut23_graphs"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(patchwork) #You need version 1.3.0 install_github("thomasp85/patchwork@v1.3.0")
library(SeuratObject)
library(Seurat)
library(EnhancedVolcano)
library(ggthemes)
library(ggprism)
library(DESeq2)
library(pheatmap)
library(EnhancedVolcano)
library(DEsingle) # From bioconductor 
library(MAST) #From bioconductor

# Weird packages for Pseudobulking 
library(SingleCellExperiment)
library(Matrix)
library(Matrix.utils) # Note this package is not available on CRAN anymore, I had to install like this: remotes::install_github("cvarrichio/Matrix.utils")
library(data.table)
```

Why can't I use Seurat 5.0? Seurat needs SeruatObject, which needs Matrix. The newest version of Matrix is only compatible with R 4.4, the compute server runs R 4.3.

### Reading in the Data

```{r}
ut23.object <- readRDS("/stor/work/Fonken/UT23_snRNAseq/local_ut23/ut23_object.rds")

# Normalize the RNA data in the ut23.object
ut23.object <- NormalizeData(ut23.object, assay = "RNA", normalization.method = "LogNormalize", scale.factor = 10000)

#Rename some things in the metadata to make my life easier when pseudobulking 
colnames(ut23.object@meta.data)[colnames(ut23.object@meta.data) == "sample"] <- "sample_id"
colnames(ut23.object@meta.data)[colnames(ut23.object@meta.data) == "group"] <- "group_id"
```

### Dimplots

```{r fig.height = 6, fig.width = 12}

# Cell Type Dot Plot
celltype_dimplot <- DimPlot(ut23.object,
        pt.size = 1,
        reduction = "umap",
        label = TRUE,
        label.size = 3,
        repel = TRUE) + 
  ggtitle("Cell Type") +
  theme_prism(base_size = 10) 
  #theme(plot.title = element_text(hjust = 0.5))  

# Experimental Group Dot Plot
sample_dimplot <- DimPlot(ut23.object,
        pt.size = 1,
        reduction = "umap",
        label = FALSE,
        label.size = 3, 
        group.by = "group_id",
        repel = TRUE) +
  ggtitle("Experimental Group") + 
  theme_prism(base_size = 10)
  #theme(plot.title = element_text(hjust = 0.5)) 

# Cell Type Expression 
markers <- list()
markers[["Microglia"]] <- c("Cx3cr1", "Csf1r", "P2ry12")
markers[["Astrocytes"]] <- c("Aqp4")
markers[["Neurons"]] <- c("Rbfox3")
markers[["Glut. Neurons"]] <- c("Slc17a7")
markers[["GABA Neurons"]] <- c("Gad1")
markers[["Oligo."]] <- c("Mog")
markers[["OPC"]] <- c("Pdgfra")
markers[["Endo."]] <- c("Pecam1")

# Create dotplot based on RNA expression
cellexpression_dotplot <- DotPlot(ut23.object, markers, assay="RNA", scale = FALSE) + 
  xlab(NULL) + 
  ylab(NULL) + 
  ggtitle("Cell Type Marker Expression") + 
  theme_prism(base_size = 10) 
  #theme(plot.title = element_text(hjust = 0.5)) 

# Patchwork plot combo 
ut23_dimplots <- (free(celltype_dimplot) + sample_dimplot) /
  cellexpression_dotplot + 
  plot_layout(heights = c(2,1))
  
ut23_dimplots + plot_annotation(tag_levels = 'A') & theme(plot.tag = element_text(size = 20))

#ggsave("/stor/work/Fonken/UT23_snRNAseq/local_ut23/Graphs/ut23_dimplots.png")
```

### Dimplot by Experimental Group

```{r fig.height = 5, fig.width = 15}
DimPlot(ut23.object,
        pt.size = 2,
        reduction = "umap",
        label = TRUE,
        label.size = 3, 
        split.by = "group_id",
        repel = TRUE)
```

### Cell Type Identification

Marker list:

Microglia:

-   Cx3cr1: receptor that binds to fralktaine, responsible for neuron to microglia communication

-   Csf1r: receptor that controls survival, proliferation and differentiation of microglia

Astrocytes:

-   Aqp4: astrocyte water channel

-   Gfap: astrocyte intermediate filament

Glut. Neurons:

-   Slc17a7: responsible for packaging glutamate into synaptic vesicles

-   Grin1: NMDA receptor subunit

GABA Neurons:

-   Gad1: Enzyme responsible for converting glutamate to GABA

-   Gabra1: subunit of GABA-A receptor

Oligodendrocytes:

-   Mog: Cell surface protein expressed in oligodendrocytes that plays a role in the integrity and function of myelin

-   Mbp: Major structural protein of myelin

OPCs:

-   Pdgfra: A receptor that regulations proliferation, migration, and differntation of OPCs into mature oligodendrocytes

Endothelial cells:

-   Pecam1: A cell adhesion molecule expressed in endothelial cells

### **Look at cell numbers by cell type**

```{r fig.height = 5, fig.width = 12}
# Compute number of cells per celltype
n_cells <- FetchData(ut23.object, 
                     vars = c("cluster_id", "group_id", "sample_id")) %>%
        dplyr::count(cluster_id, group_id, sample_id)

# Barplot of number of cells per celltype by group
ggplot(n_cells, aes(x=cluster_id, y=n, fill=group_id)) +
    geom_bar(position=position_dodge(), stat="identity") +
    geom_text(aes(label=n), vjust = -.2, position=position_dodge(1))

# Barplot of number of cells per celltype by sample
ggplot(n_cells, aes(x=cluster_id, y=n, fill=sample_id)) +
    geom_bar(position=position_dodge(), stat="identity") +
    geom_text(aes(label=n), vjust = -.2, position=position_dodge(1))

# Barplot of number of cells per celltype by sample
ggplot(n_cells, aes(x=cluster_id, y=n, fill=sample_id)) +
    geom_bar(position=position_dodge(), stat="identity") +
    geom_text(aes(label=n), vjust = -.2, position=position_dodge(1))
```

### Differential Expression Analysis on Bulk Data

```{r}
# pseudobulk the counts based on donor-condition-celltype
bulk.ut23.all <- AggregateExpression(ut23.object, 
                                   assays = "RNA", 
                                  slot = "counts",
                                   return.seurat = T, 
                                   group.by = c("group_id", "sample_id"))

# Adding the additional metadata column (this was the easiest way I could think to do this, probably a better way)
group_ids <- substr(bulk.ut23@meta.data$orig.ident, 1, 2)
bulk.ut23[["group_id"]] <- group_ids

# This is making a metadata file for the idents: 
bulk.ut23$celltype.stim <- paste(bulk.ut23$group_id)

# Setting the idents
Idents(bulk.ut23) <- "group_id"

# Run FindMarkers
ut23.findmarkers <- FindMarkers(object = bulk.ut23,
                                assay = "RNA",
                                slot = 'counts',
                         ident.1 = "YV",
                         ident.2 = "AV",
                         test.use = "DESeq2")

# Volcano Plots 
EnhancedVolcano(ut23.findmarkers,
                lab = rownames(ut23.findmarkers),
                x = "avg_log2FC",
                y = "p_val_adj",
                FCcutoff = 0.25,
                pCutoff = 0.05,
                title = "Young Vehicle vs Aged Vehicle")

ggsave("/stor/work/Fonken/UT23_snRNAseq/local_ut23/Graphs/YVvsAV.png", width = 10, height = 10)
```

### Differential Expression Analysis With Pseudobulking

Trying this tutorial, skipped all the QC steps since I already did those

<https://www.youtube.com/watch?v=04gB2owLKus>

Also followed this tutorial:\
<https://hbctraining.github.io/scRNA-seq_online/lessons/pseudobulk_DESeq2_scrnaseq.html>

```{r}
# pseudobulk the counts based on donor-condition-celltype
bulk.ut23 <- AggregateExpression(ut23.object, 
                                   assays = "RNA", 
                                  slot = "counts",
                                   return.seurat = T, 
                                   group.by = c("group_id", "sample_id", "cluster_id"))

# Adding in two additional metadata columns (this was the easiest way I could think to do this, probably a better way)
cluster_ids <- sub(".*_", "", rownames(bulk.ut23@meta.data))
bulk.ut23@meta.data[["cluster_id"]] <- cluster_ids

group_ids <- substr(bulk.ut23@meta.data$orig.ident, 1, 2)
bulk.ut23[["group_id"]] <- group_ids

# This is making a metadata file for the idents: 
bulk.ut23$celltype.stim <- paste(bulk.ut23$cluster_id, bulk.ut23$group_id, sep = "_")

# Setting the idents
Idents(bulk.ut23) <- "celltype.stim"

# Run FindMarkers
ut23.findmarkers <- FindMarkers(object = bulk.ut23,
                                assay = "RNA",
                                slot = 'counts',
                         ident.1 = "Astrocytes_YV",
                         ident.2 = "Astrocytes_AV",
                         test.use = "DESeq2")

# Run FindMarkers
ut23.findmarkers <- FindMarkers(object = bulk.ut23,
                                assay = "RNA",
                                slot = 'data',
                         ident.1 = "Astrocytes_YV",
                         ident.2 = "Astrocytes_AV",
                         test.use = "MAST")

# Volcano Plots 
EnhancedVolcano(ut23.findmarkers,
                lab = rownames(ut23.findmarkers),
                x = "avg_log2FC",
                y = "p_val_adj",
                FCcutoff = 0.25,
                pCutoff = 0.05,
                title = "Astrocytes: Young Vehicle vs Aged Vehicle")

# "Astrocytes: Young Vehicle vs Aged Vehicle"
# "Astrocytes: Aged M Vaccae vs Aged Vehicle"

ggsave("/stor/work/Fonken/UT23_snRNAseq/local_ut23/Graphs/astro_YVvsAV.png", width = 10, height = 10)
```

### Trying out MAST

<https://miaozhun.github.io/DEsingle/>

No need for pseudobulking, trying to use findmarkers() first

```{r}
# Subsetting microglia cluster 
microglia_subcluster <- subset(ut23.object, idents = "Microglia")

# Setting default assay to RNA 
DefaultAssay(microglia_subcluster) <- "RNA"

# Set the identity class to the group metadata
Idents(microglia_subcluster) <- microglia_subcluster$group_id

# Run FindMarkers using MAST to compare AV vs YV
microglia_mast_yvav <- FindMarkers(microglia_subcluster, 
                            ident.1 = "YV", 
                            ident.2 = "AV", 
                            test.use = "MAST")

# Run FindMarkers using MAST to compare AV vs YV
microglia_mast_amav <- FindMarkers(microglia_subcluster, 
                            ident.1 = "AM", 
                            ident.2 = "AV", 
                            test.use = "MAST")

# Volcano Plots 
EnhancedVolcano(microglia_mast_yvav,
                lab = rownames(microglia_mast_yvav),
                x = "avg_log2FC",
                y = "p_val_adj",
                FCcutoff = 0.25,
                pCutoff = 0.05,
                title = "Young Vehicle vs Aged Vehicle")

ggsave("/stor/work/Fonken/UT23_snRNAseq/local_ut23/Graphs/astro_YVvsAV.png", width = 10, height = 10)

# Volcano Plots 
EnhancedVolcano(microglia_mast_amav,
                lab = rownames(microglia_mast_amav),
                x = "avg_log2FC",
                y = "p_val_adj",
                FCcutoff = 0.25,
                pCutoff = 0.05,
                title = "Aged M Vaccae vs Aged Vehicle")

ggsave("/stor/work/Fonken/UT23_snRNAseq/local_ut23/Graphs/astro_AVvsAM.png", width = 10, height = 10)

#creating deg lists
genes_microglia_mast_yvav <- rownames(microglia_mast_yvav[microglia_mast_yvav$p_val_adj < 0.05, ])

write.table(genes_microglia_mast_yvav, file = "significant_genes.txt", quote = FALSE, row.names = FALSE, col.names = FALSE)



```

### Pseudobulking (1 of 2)

Was trying this one: <https://github.com/hbctraining/scRNA-seq_online/blob/master/lessons/09_merged_SC_marker_identification.md>

# I need to add batch here

```{r fig.height = 10, fig.width = 10}

# Extract raw counts and metadata to create SingleCellExperiment object
counts <- ut23.object@assays$RNA@counts
metadata <- ut23.object@meta.data

# Set up metadata as desired for aggregation and DE analysis
metadata$cluster_id <- factor(ut23.object@active.ident)

sce <- SingleCellExperiment(assays = list(counts = counts), 
                           colData = metadata)

# Extract unique names of clusters (= levels of cluster_id factor variable), I have 7 clusters 
cluster_names <- levels(colData(sce)$cluster_id)
cluster_names

# Extract unique names of samples (= levels of sample_id factor variable)
# Factoring first 
colData(sce)$sample_id <- factor(colData(sce)$sample_id)

sample_names <- levels(colData(sce)$sample_id)
sample_names

# Total number of samples (we have 9 samples )
length(sample_names)

# Subset metadata to include only the variables you want to aggregate across (here, we want to aggregate by sample and by cluster)
groups <- colData(sce)[, c("cluster_id", "sample_id")]
head(groups)

# Aggregate across cluster-sample groups
# transposing row/columns to have cell_ids as row names matching those of groups
aggr_counts <- aggregate.Matrix(t(counts(sce)), 
                                groupings = groups, fun = "sum") 

# Explore output matrix
class(aggr_counts)
dim(aggr_counts)
aggr_counts[1:6, 1:6]

# Transpose aggregated matrix to have genes as rows and samples as columns
aggr_counts <- t(aggr_counts)
aggr_counts[1:6, 1:6]

# Understanding tstrsplit()

## Exploring structure of function output (list)
tstrsplit(colnames(aggr_counts), "_") %>% str()

## Comparing the first 10 elements of our input and output strings
head(colnames(aggr_counts), n = 10)
head(tstrsplit(colnames(aggr_counts), "_")[[1]], n = 10)

# Using which() to look up tstrsplit() output
microglia_idx <- which(tstrsplit(colnames(aggr_counts), "_")[[1]] == "Microglia")
microglia_idx

colnames(aggr_counts)[microglia_idx]
aggr_counts[1:10, microglia_idx]

# As a reminder, we stored our cell types in a vector called cluster_names
cluster_names


# Loop over all cell types to extract corresponding counts, and store information in a list

## Initiate empty list
counts_ls <- list()

for (i in 1:length(cluster_names)) {

  ## Extract indexes of columns in the global matrix that match a given cluster
  column_idx <- which(tstrsplit(colnames(aggr_counts), "_")[[1]] == cluster_names[i])
  
  ## Store corresponding sub-matrix as one element of a list
  counts_ls[[i]] <- aggr_counts[, column_idx]
  names(counts_ls)[i] <- cluster_names[i]

}

# Explore the different components of the list
str(counts_ls)

# Reminder: explore structure of metadata
head(colData(sce))

# Extract sample-level variables
metadata <- colData(sce) %>% 
  as.data.frame() %>% 
  dplyr::select(group_id, sample_id)

dim(metadata)
head(metadata)

# Exclude duplicated rows
metadata <- metadata[!duplicated(metadata), ]

dim(metadata)
head(metadata)

# Rename rows
rownames(metadata) <- metadata$sample_id
head(metadata)

# Number of cells per sample and cluster
t <- table(colData(sce)$sample_id,
           colData(sce)$cluster_id)
t[1:6, 1:6]

# Creating metadata list-----------

## Initiate empty list
metadata_ls <- list()

for (i in 1:length(counts_ls)) {
  
    ## Initiate a data frame for cluster i with one row per sample (matching column names in the counts matrix)
    df <- data.frame(cluster_sample_id = colnames(counts_ls[[i]]))
    
    ## Use tstrsplit() to separate cluster (cell type) and sample IDs
    df$cluster_id <- tstrsplit(df$cluster_sample_id, "_")[[1]]
    df$sample_id  <- tstrsplit(df$cluster_sample_id, "_")[[2]]
    
    
    ## Retrieve cell count information for this cluster from global cell count table
    idx <- which(colnames(t) == unique(df$cluster_id))
    cell_counts <- t[, idx]
    
    ## Remove samples with zero cell contributing to the cluster
    cell_counts <- cell_counts[cell_counts > 0]
    
    ## Match order of cell_counts and sample_ids
    sample_order <- match(df$sample_id, names(cell_counts))
    cell_counts <- cell_counts[sample_order]
    
    ## Append cell_counts to data frame
    df$cell_count <- cell_counts
    
    
    ## Join data frame (capturing metadata specific to cluster) to generic metadata
    df <- plyr::join(df, metadata, 
                     by = intersect(names(df), names(metadata)))
    
    ## Update rownames of metadata to match colnames of count matrix, as needed later for DE
    rownames(df) <- df$cluster_sample_id
    
    ## Store complete metadata for cluster i in list
    metadata_ls[[i]] <- df
    names(metadata_ls)[i] <- unique(df$cluster_id)

}

# Explore the different components of the list
str(metadata_ls)
```

**DESeq2 on Pseudobulked data (2 of 2)**

```{r}
# Double-check that both lists have same names
all(names(counts_ls) == names(metadata_ls))

# Focusing on Microglia 
idx <- which(names(counts_ls) == "Microglia")
cluster_counts <- counts_ls[[idx]]
cluster_metadata <- metadata_ls[[idx]]

# Check contents of extracted objects
cluster_counts[1:6, 1:6]
head(cluster_metadata)

# Check matching of matrix columns and metadata rows
all(colnames(cluster_counts) == rownames(cluster_metadata))

# Setting the reference level as AV
cluster_metadata$group_id <- factor(cluster_metadata$group_id, levels = c("AV", "YV", "AM"))

# Create DESeq2 object        
dds <- DESeqDataSetFromMatrix(cluster_counts, 
                              colData = cluster_metadata, 
                              design = ~ batch + group_id)

# Transform counts for data visualization
rld <- rlog(dds, blind=TRUE)

# Plot PCA
DESeq2::plotPCA(rld, ntop = 500, intgroup = "group_id")

#plotting PCA based on cell count 
DESeq2::plotPCA(rld, ntop = 500, intgroup = "cell_count")

# Extract the rlog matrix from the object and compute pairwise correlation values
rld_mat <- assay(rld)
rld_cor <- cor(rld_mat)

# Plot heatmap
pheatmap(rld_cor, annotation = cluster_metadata[, c("group_id"), drop=F])

# PCA doesn't plot neatly, and I'm concerned about cell count clustering 

# Run DESeq2 differential expression analysis (with glmGamPoi and LRT)
# https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#likelihood-ratio-test

dds <- DESeq(dds,
             fitType = "glmGamPoi",
             test = "LRT",
             reduced = ~1)

# Plot dispersion estimates
plotDispEsts(dds)

# dispersion plot looks encouraging, since we expect our dispersions to decrease with increasing mean and follow the line of best fit (in red).

# Check the coefficients for the comparison
resultsNames(dds)

# ******Sommething is wrong here- I need to figure out how to change the contrasts to get the correct group_id combinations

# Generate results object
res <- results(dds, 
               name = "group_idAM",
               alpha = 0.05)

# Shrink the log2 fold changes to be more appropriate using the apeglm method - should cite [paper]() when using this method
res <- lfcShrink(dds, 
                 coef = "group_idYV",
                 res=res,
                 type = "apeglm")

# Turn the DESeq2 results object into a tibble for use with tidyverse functions
res_tbl <- res %>%
  data.frame() %>%
  rownames_to_column(var = "gene") %>%
  as_tibble() %>%
  arrange(padj)

# Check results output
res_tbl 

# Set thresholds
padj_cutoff <- 0.05

# Subset the significant results
sig_res <- dplyr::filter(res_tbl, padj < padj_cutoff) %>%
  dplyr::arrange(padj)

# Check significant genes output
sig_res
```

### Clustering the Microglia Cluster

There is a function canned FindSubCluster but I don't think it actually reclusters the data, there aren't a lot of resources online about how to do this but I'm roughly following this:\
<https://github.com/satijalab/seurat/issues/4320>\
\
There seem to be several ways to do it and I'm basically just totally re-doing the clustering with the subsetted object.

```{r}
# Subsetting one cluster
microglia_subcluster <- subset(ut23.object, idents = "Microglia")

# Changing the default assay from SCT to RNA
DefaultAssay(microglia_subcluster) <- "RNA"

#  Going through the guided-clustering tutorial's workflow

# Run Split Object
split_microglia_subcluster <- SplitObject(microglia_subcluster, split.by = "group_id")
 
# SCTransform for each sample 
for (i in 1:length(split_microglia_subcluster)) {
    split_microglia_subcluster[[i]] <- SCTransform(split_microglia_subcluster[[i]], vst.flavor = "v2")
}

# Select the most variable features to use for integration
integ_features <- SelectIntegrationFeatures(object.list = split_microglia_subcluster, 
                                            nfeatures = 3000) 

# Prepare the SCT list object for integration
split_microglia_subcluster <- PrepSCTIntegration(object.list = split_microglia_subcluster, 
                                   anchor.features = integ_features)

# Find best buddies - can take a while to run
integ_anchors <- FindIntegrationAnchors(object.list = split_microglia_subcluster, 
                                        normalization.method = "SCT", 
                                        anchor.features = integ_features)

# Integrate across conditions
int_microglia <- IntegrateData(anchorset = integ_anchors, 
                                   normalization.method = "SCT")

# Run RunPCA (Don't need to do FindVariableFeatures or ScaleData because we're using SCTransform)
int_microglia <- RunPCA(int_microglia)

# Elbow plot to identify dimensionality
ElbowPlot(int_microglia)

# Use the elbow from above to subcluster the cells
int_microglia <- FindNeighbors(int_microglia, dims = 1:10)
int_microglia <- FindClusters(int_microglia, resolution = 0.2)
int_microglia <- RunUMAP(int_microglia, dims = 1:10)

#UMAP Plot of the clusters by condition  
DimPlot(int_microglia, 
        reduction = "umap", 
        pt.size = 1,
        split.by = "group_id", 
        label = TRUE) 
  
# Assign identity of clusters
#Set the cell type in the metadata 
int_microglia$microglia.subcluster <- Idents(int_microglia)

```

```{r fig.height = 5, fig.width = 12}
# Compute number of cells per celltype
n_cells_micro <- FetchData(int_microglia, 
                     vars = c("microglia.subcluster", "group_id")) %>%
        dplyr::count(microglia.subcluster, group_id)

# Barplot of number of cells per celltype by sample
ggplot(n_cells_micro, aes(x=microglia.subcluster, y=n, fill=group_id)) +
    geom_bar(position=position_dodge(), stat="identity") +
    geom_text(aes(label=n), vjust = -.2, position=position_dodge(1))
```

```{r}
# Finding markers in cluster 1, where we see the change we are interested in 
cluster1.markers <- FindMarkers(int_microglia, ident.1 = 1)

# Finding markers for each cluster compared to the other clusters 
cluster.markers <- FindAllMarkers(int_microglia, only.pos = TRUE)
cluster.markers %>%
    group_by(cluster) %>%
    filter(cluster == "1")

# Feature Plots loooking at genes that may be altered in aging 
genelist <- c("Cmtm4")

DotPlot(object = int_microglia, features = genelist, col.min=0) 
```

\
